---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# L1fdvGenetics

<!-- badges: start -->
<!-- badges: end -->

The goal of `L1fdvGenetics` is to provide interactive tutorials for the Genetics project. Tutorials are packaged so you can install them on you computer and do the exercices without access to an internet connection. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("mahendra-mariadassou/L1fdvGenetics")
```

<!-- If the installation of `devtools` fails, you may need to install some system dependencies (for example on ubuntu/debian): -->
<!-- ```{bash, eval = FALSE} -->
<!-- sudo apt-get install libxml2-dev libcurl4-openssl-dev libssl-dev -->
<!-- ``` -->

<!-- On Mac, you may need some of the compiling tools listed [here](https://cran.r-project.org/bin/macosx/tools/) -->

On **Windows**: you may need **Rtools** and **git** 

- **Rtools**: visit the dedicated [page](https://cran.r-project.org/bin/windows/Rtools/), download the suggested exe and install it on your computer
- **git**: visit the dedicated [page](https://git-scm.com/download/win), download the suggested exe and install it on your computer

## Starting a tutorial

This package is intended for use with `learnr` (which should be installed when you install `L1fdvGenetics`)

```{r example, eval = FALSE}
library(learnr)
```

## Candidate gene (single marker) approach

```{r simple_marker, eval = FALSE}
## Launch only one tutorial at the time!!
learnr::run_tutorial("01_candidate_gene", package = "learningr")
```

## QTL principle

```{r QTL, eval = FALSE}
## Launch only one tutorial at the time!!
learnr::run_tutorial("02_multiple_projects", package = "learningr")
```

## Simple Interval Mapping (SIM)

```{r sim, eval = FALSE}
## Launch only one tutorial at the time!!
learnr::run_tutorial("03_interval_ammping", package = "learningr")
```
